---
title: "Australia monthly CPI indicator"
author: "Nick Twort"
date: "`r format(Sys.time(), '%d %B %Y')`"
format: html
execute: 
  echo: false
  warning: false
---

## The monthly index

```{r}
library(tidyverse)
library(here)
library(gt)
df <- read_csv(here("data/raw/cpi_weights.csv"))
source(here("source/Deeper dive.R")) 
```


@fig-all-groups below shows that the [monthly CPI indicator]{style="color: #008698; font-weight: bold"} tracks closely with the [existing quarterly indicator]{style="color: #232C31; font-weight: bold"}. Inflation for the year to `r monthly_cpi |> pull(date) |> max() |> format("%B %Y")` was `r monthly_cpi |> filter(series_id == "A128478317T") |> tail(1) |> pull(value_ch) |> scales::percent(accuracy = 0.1)`, according to the monthly indicator.


```{r}
#| label: fig-all-groups
#| fig-cap-location: margin
#| fig-cap: Different measures of household inflation
#| fig-width: 8
#| fig-height: 3.5

series <- c("A2325846C", # quarterly all groups, original
            "A128478317T" # all groups CPI, original
) 

create_full_chart(series = c("A2325846C", "A128478317T"))


```


The monthly index shows that the `r lookup_series_name(get_nth_change_yoy())` group increased most over the last year -- by `r monthly_cpi |> filter(series_id == get_nth_change_yoy()) |> filter(date == max(date)) |> pull(value_ch) |> scales::percent(accuracy = 0.1)`, shown in @fig-1.


```{r}
#| label: fig-1
#| fig-cap-location: margin
#| fig-cap: !expr paste0(lookup_series_name(get_nth_change_yoy()), " prices")
#| fig-width: 8
#| fig-height: 3.5

n <- 1
individual_chart(get_nth_change_yoy(n), n)


```

The `r lookup_series_name(get_nth_change_yoy(2))` group increased the next most -- by `r monthly_cpi |> filter(series_id == get_nth_change_yoy(2)) |> filter(date == max(date)) |> pull(value_ch) |> scales::percent(accuracy = 0.1)` -- see @fig-2.


```{r}
#| label: fig-2
#| fig-cap-location: margin
#| fig-cap: !expr paste0(lookup_series_name(get_nth_change_yoy(2)), " prices")
#| fig-width: 8
#| fig-height: 3.5

n <- 2
individual_chart(get_nth_change_yoy(n), n)


```
A summary of changes at the group level is show in @tbl-ch below:

```{r}
#| label: tbl-ch
#| tbl-cap-location: margin
#| tbl-cap: CPI change by group
print_latest_by_group() |> 
  rename(Group = series_2, `YoY change` = value_ch) |> 
  gt() |> 
  fmt_percent(columns = 2, decimals = 1) |> 
  tab_options(column_labels.background.color = "#232C31")


```
